A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
Restate cyclicity/bisimilarity rules, cleanup.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t : \tau, \tau \in \Data
  \quad
  p \in \ttpath{\Ec}{\ec{t}}{\ec{t}}
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau, \tau \in \Codata
  \quad
 \exists \sigma. p \sigma \in \tpath{\Ec}{\ec{t_1}} \text{ for each } p \in \tpath{\Ec}{\ec{t_2}}
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity.
}
\label{fig:ab-rules}
\end{figure}

@@

To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labeled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
In other words, $t_1$ is equivalent to a $C$-application whose $n^{th}$ argument is equivalent to $t_2$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the \emph{source} of a path is the source of its first edge, and the \emph{destination} of a path is the destination of its last edge.
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.
We will write, for instance, $\ec{t_1} \rightarrow^{C_1}_{n_1} \ec{t_2} \rightarrow^{C_2}_{n_2} \ec{t_3}$ 
as shorthand for the path containing an edge from $\ec{t_1}$ to $\ec{t_2}$, followed by one from $\ec{t_2}$ to $\ec{t_3}$.

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $p \in \ttpath{\Ec}{\ec{t}}{\ec{t}}$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In particular, if there is a mapping $\sigma$ over equivalence classes such that all paths $p$ from $t_2$
have a corresponding path $p \sigma$ from $t_1$,
then $t_1$ must be equal to $t_2$.
We demonstrate this rule concretely with the following example.

\begin{example}
Let $\Ec$ be the set $\{ x \teq \const{LCons}(y,\const{LCons}(y,x)), x \tneq \const{LCons}(y,x) \}$.  \rem{ensure that \const{LCons} understood}
The equivalence classes of $\Ec$ are $\{ x, \const{LCons}(y,\const{LCons}(y,x)) \}$, $\{ \const{LCons}(y,x) \}$, and $\{ y \}$.
The set $\tpath{\Ec}{\ec{x}}$ is:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{x} \rightarrow^{\const{LCons}}_1 & \ec{y}, & & \\ 
   & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_1 & \ec{y},& \\
   & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} & \}.
\end{array}\]
The set $\tpath{\Ec}{\ec{\const{LCons}(y,x)}}$ is:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_1 & \ec{y}, & & \\ 
   & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} \rightarrow^{\const{LCons}}_1 & \ec{y},& \\
   & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} & \}.
\end{array}\]
Thus, we may conclude $x \teq \const{LCons}(y,x)$ by the rule \rn{Bisimilar}
for the mapping $\sigma :=$ $\{ \ec{x} \mapsto \ec{\const{LCons}(y,x)}, \ec{\const{LCons}(y,x)} \mapsto \ec{x} \}$,
after which the rule \rn{Conflict} may be applied.
\end{example}

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, t : \tau \text{ finite}, s \in \Sel^\tau  \}$.
Let $S^\ast$ be the fixed point of this sequence, which is guarenteed to be a finite set 
since $S^i \setminus S^0$ only includes selectors applied to terms having finite type.
Now, let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_j}_i( t ) ) \mid t \in S^\ast, t : \tau, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
obtained as a result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the maximum depth of a branch in a derivation tree with root node $\Ec$ is $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail exactly one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, or \rn{Bisimilar}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and since $\Ec \models_{\thD} t_1 \teq t_2$,
we have that $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

>>>>

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t : \tau
  \quad
  \tau \in \Data
  \quad
  \ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau
  \quad
 \tau \in \Codata
  \quad
 \tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity.
}
\label{fig:ab-rules}
\end{figure}

@@

To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labeled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
In other words, $t_1$ is equivalent to a $C$-application whose $n^{th}$ argument is equivalent to $t_2$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the \emph{source} of a path is the source of its first edge, and the \emph{destination} of a path is the destination of its last edge.
We say a path $p$ is \emph{maximal} if its destination $\ec{t}$ is either
the source of an edge in $p$ or does not contain an application of a constructor (including the case when $t$ is not of datatype/codatatype type).
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all maximal paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.
We will write, for instance, $\ec{t_1} \rightarrow^{C_1}_{n_1} \ec{t_2} \rightarrow^{C_2}_{n_2} \ec{t_3}$ 
as shorthand for the path containing an edge from $\ec{t_1}$ to $\ec{t_2}$, followed by one from $\ec{t_2}$ to $\ec{t_3}$.

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In particular, if there is a mapping $\sigma$ over equivalence classes such that all paths $p$ from $t_2$
have a corresponding path $p \sigma$ from $t_1$,
then $t_1$ must be equal to $t_2$.
We demonstrate this rule concretely with the following example.
\rem{ensure that \const{LCons} understood}

\begin{example}
Let $\Ec$ be the set $\{ x \teq \const{LCons}(y,\const{LCons}(y,x)) \}$.
After taking the bidirectional closure of $\Ec$, we obtain equivalence classes 
$\{ x, \const{LCons}(y,\const{LCons}(y,x)) \}$, $\{ \const{LCons}(y,x) \}$, and $\{ y \}$.
The set $\tpath{\Ec}{\ec{x}}$ is:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{x} \rightarrow^{\const{LCons}}_1 & \ec{y}, & & \\ 
   & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_1 & \ec{y},& \\
   & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} & \}.
\end{array}\]
The set $\tpath{\Ec}{\ec{\const{LCons}(y,x)}}$ is:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_1 & \ec{y}, & & \\ 
   & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} \rightarrow^{\const{LCons}}_1 & \ec{y},& \\
   & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} & \}.
\end{array}\]
Thus, we may infer $x \teq \const{LCons}(y,x)$ by the rule \rn{Bisimilar}
for the mapping $\sigma :=$ $\{ \ec{x} \mapsto \ec{\const{LCons}(y,x)}, \ec{\const{LCons}(y,x)} \mapsto \ec{x} \}$.
\qed
\end{example}

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, t : \tau \text{ finite}, s \in \Sel^\tau  \}$.
Let $S^\ast$ be the fixed point of this sequence, which is guarenteed to be a finite set 
since $S^i \setminus S^0$ only includes selectors applied to terms having finite type.
Now, let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t \in S^\ast, t : \tau, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
the result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail exactly one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, or \rn{Bisimilar}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}
